---
title: "Analysis of longitudal data"
author: "Savi Virolainen"
date: "11/30/2020"
output: 
  html_document:
    theme: darkly
    highlight: breezedark
---

# Analysis of longitudal data

## Implementing the analyses of Chapter 8 of MABS using RATS data
Load the data:
```{r}
RATS <- read.csv("data/RATS.csv")
RATSL <- read.csv("data/RATSL.csv")

# ID and Group as factors (they were transformed back to int when reading): 
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
```

### Introduction
Table of wide RATS data:
```{r}
head(RATS, n=nrow(RATS))
```

### Graphical displays of longitudal data
The RATS data contains measured body weights of rats in different points of time where the rats are divided into three groups. For each of the three group, different diets were fed to the rats. We examine the data graphically where the three groups are presented in different figures where the body weight is in the vertical axis and time in the horizontal axis:
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

Two kind of patterns are apparent for all groups: the weight seems to increase as a function of time and rats with higher starting weight generally have higher weight throughout the study. This is generally referred to as tracking. However, if the starting weights are close to each other, the ordering of the final weights seems to be often different than the ordering of the starting weights.

We then do the same graphical examination but with standardized data (=centered and divided by standard deviation):
```{r}
# Create standardized weights
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Plot the standardized weights
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdweight), max(RATSL$stdweight)))
```

One observe the tracking phenomenom from the standardized data summary above as well. 

### Summary measure analysis of longitudal data

We draw summary figure and plot the mean profiles:
```{r, message=FALSE}
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(length(unique(RATSL$Time)))) %>%
  ungroup()

ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1, 2, 3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1, 2, 3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9, 0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

The weights seem to be generally increasing in time, the slowest in group 1, the next slowest in group 3, and the fastest in group 2. Also, there seems to be least variation in weights within the first group, a bit more in the third group, while in group 2 there is most variation in the weights. One standard error bounds don't cross in any of the groups so the mean weights seem to generally differ between the groups at all time points.  

Then we draw another kind of summary figure, now using boxplots for the weights:
```{r, message=FALSE, warning=FALSE}
ggplot(RATSL, aes(x = factor(Time), y = Weight, fill=Group)) +
  geom_boxplot()
```

One may obserce presence of some outliers in groups 1 and 3 but no outliers in group 2. The weights seem to be generally increasing, with the least within group variation in group 1, and little within group variation in group 3 as well but more within group variation in group 2. Because the medians are generally not in the middle, the weight distributions seem skewed.

We draw the boxplots of mean versus group. We remove the initial/baseline time point as in the book.
```{r, message=FALSE, warning=FALSE}
# Remove the initial time point:
RATSS2 <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight)) %>%
  ungroup()

# Function to draw a boxplot:
draw_bp <- function(dat) {
  ggplot(dat, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time > 1")
}

# Boxplot:
draw_bp(RATSS2)
```

It seems the the first group has smaller means than the other two groups. The difference is smaller between groups 2 and 3 but still the means of the second group seem significantly smaller than the one of the third group, as besides a single outlier, the group 2's mean weights are smaller than group 3's mean weights. 

Following the book, we remove outlier in group 2 that may significantly alter the results, and draw the boxplot again:
```{r}
RATSS3 <- filter(RATSS2, mean < 550)
draw_bp(RATSS3)
```
Now the difference between the first and third groups is more apparent.

We formally test the group difference using two-sample t-tests without the outlier:
```{r}
t.test(mean ~ Group, data = filter(RATSS3, Group %in% c(1, 2)), var.equal=FALSE) # Groups 1 and 2
t.test(mean ~ Group, data = filter(RATSS3, Group %in% c(1, 3)), var.equal=FALSE) # Groups 1 and 3
t.test(mean ~ Group, data = filter(RATSS3, Group %in% c(2, 3)), var.equal=FALSE) # Groups 2 and 3
```

The small p-values indicate that the mean weights of all the groups are most likely different. The above printout also shows the means of the different groups and confidence intervals the difference of the means. 

Then we run analysis of variance for the model. First, we add back the baseline time point to the data (that includes the outlier).
```{r}
RATSS4 <- mutate(RATSS2, baseline=RATS$WD1)
fit <- lm(mean ~ baseline + Group, data=RATSS4)
anova(fit)
```

As the above summary printout of anova shows, the baseline Weight is strongly related to the Weights recorded after following the diet (as was also concluded from the graphical summaries already). The group has some (but very strong) statistical significance as well, indicating that the different diets may as well be related to the different weights recorded. Recall that the weight gaining speeds were noted to differ a bit between the groups in the earlier graphical summaries, being in line with the results from anova.

We don't have missing values in the data, so that part of the book's chapter 8 wont be replicated here. 

## Implementing the analyses of Chapter 9 of MABS using BPRS data
Load the data:
```{r}
BPRS <- read.csv("data/BPRS.csv")
BPRSL <- read.csv("data/BPRSL.csv")

# Treatment, subject, and ID as factors (they were transformed back to int when reading):
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
BPRSL$ID <- factor(BPRSL$ID)
```

### How Do Rats Grow? (with BPRS data)
Table of wide form BPRS data:
```{r}
head(BPRS, n=nrow(BPRS))
```

Table of long from BPRS data for the baseline week;
```{r}
head(BPRSL, n=40)
```

Plot of bprs against week (time) for BPRS data, ignoring the repeated-measures structure of the data:
```{r}
plot(bprs ~ week, data=BPRSL, pch=c("1", "2")[treatment])
```

Then we create a graphical display of the data that takes into account the longitudal structure:
```{r}
# NOTE THAT WE CREATED THE ID VARIABLE IN THE WRANGLING PART
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group=ID)) +
        geom_line(aes(color=treatment, linetype=treatment)) +
        scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, by=2)) +
        scale_y_continuous(name = "bprs") +
        theme(legend.position = "top")
p1
```
the above figure shows the individual subjects bprs profiles. The bprs seems to be generally decreasing slightly but it is difficult to detect any apparent differences between the two treatments.


Scatter plot matrix of repreated measures in BPRS data:
```{r}
par(las=1)
pairs(BPRS[,3:11], gap=0.4, font.labels=2, cex.labels=1, panel=function(x, y) {
  points(x, y, col="blue", pch="*", cex=1.5)
  abline(lm(y ~ x), col="red")
  },
  lower.panel=function(x, y) {
    old_par <- par(no.readonly=TRUE)
    on.exit(old_par)
    par(usr=c(0, 1, 0, 1))
    ncol <- 10
    colpal <- colorRampPalette(c("skyblue", "blue", "red", "red4"))(ncol)
    corcol <- abs(round(ncol*cor(x, y)))
    do.call(rect, c(col=colpal[corcol], list(0, 0, 1, 1)))
    text(x=0.5, y=0.5, labels=round(cor(x, y), 2), font=ifelse(abs(cor(x, y)) > 0.7, 4, 1),
         cex=1.2)
  })
```
Due to the linear patterns in the scatterplots in the upper-right triangle above, it seems that the repeated measurements are far from independent from each other, making the independence assumption in the OLS regression inappropriate (although I think OLS estimator generally has good properties under much weaker assumptions than independence). The lower-right triangle show the sample correlations with high red color indicating high correlation - this also shows strong linear depedence between the weeks.

We fit linear regression model via OLS to the BPRS data with bprs as response, and treatment and week (time) as explanatory variables, and ignoring the repeated measures structure of the data:
```{r}
summary(lm(bprs ~ treatment + week, data=BPRSL))
```
According to the t-tests, treatment does not seem to have statistically significant effect on the response but the time variable (week) definitely does.

Next, we fit the random intercepts model and examine the summary printout:
```{r, message=FALSE}
library(lme4)
fit1 <- lmer(bprs ~ week + treatment + (1 | ID), data = BPRSL, REML = FALSE)
summary(fit1)
```
Focusing on the fixed effects, the small t-value of "treatment2" indicates that there is no statistically significant difference in the intercepts for observations with different treatments. The week variable with negative coefficient has large t-value which implies that the bprs seems to generally decrease as the time passes in the experiment, as was concluded from the grapchical summary as well. Focusing then on the random effects, the variance and standard deviations of the random effect intercepts is quite large, which tells that the individuals seem to have quite different intercepts (=starting bprs scores) in general. 

Then we fit a random intercept and slope model with week and treatment as explanatory variables and examine the summary printout:
```{r}
fit2 <- lmer(bprs ~ week + treatment + (week | ID), data = BPRSL, REML = FALSE)
summary(fit2)
```
Focusing on the random effects, notice that the variance of the random intercepts increased while the variance for the random slope also seems relative large. Focusing then on the fixed effects, the week parameter with negative coefficient estimate has a large t-value, so also in this model the bprs score seems to decrease over time. The t-value for the treatment coefficient is still small, indicating that there is no statistically significant difference between in the intercepts between the two treatments. 

Next, we estimate a random intercept and random slope model that also allows for interactions between the treatment and week variable. That is, subjects with different treatments may have different fixed effect slopes. We also examine the summary printout:
```{r}
fit3 <- lmer(bprs ~ week*treatment + (week | ID), data = BPRSL, REML = FALSE)
summary(fit3)
```
Focusing on the random effects, the variances of the random intercepts and slopes remained approximately the same compared to the model without interactions. Focusing then on the fixed effects, the week variable still has statistically significant negative coefficient. The treatment2-intercept, however, now became negative when it was positive before, but its t-value is still too close to zero to declare any statistical significance. The interaction between week and treatment has t-value 1.349 which implies the approximate p-value
```{r}
2*(1 - pnorm(1.349))
```
for the null hypothesis that the interaction is zero. The interaction is more statistically significant than the intercept difference of treatments (which should be included to control for different starting scores) but due to the large p-value we cannot say with great certainty that the bprs scores decrease with different speed for subjects with different treatments. The point estimate, nonetheless, implies that the bprs score decreases less over time for subjects with treatment 2 than with treatment 1. 

After examining the summary printout, we visualize the model with random intercepts and slopes as well as fixed effects interaction for treatment and time slope. We also plot the observed values next to the fitted values.
```{r, message=FALSE}
library(ggplot2)
library(ggpubr)
BPRSL <- mutate(BPRSL, Fitted=fitted(fit3)) # Add the fitted values to the data

p2 <- ggplot(BPRSL, aes(x = week, y = Fitted, group = ID)) +
        geom_line(aes(linetype = treatment, color=treatment)) +
        scale_x_continuous(name = "Time (week)", breaks = seq(0, 8, by=2)) +
        scale_y_continuous(name = "bprs") +
        theme(legend.position = "top")
ggarrange(p1, p2, labels=c("Observed", "Fitted"), ncol=2, nrow=1)
```

Looking at the fitted values, the conclusions from the summary printout are apparent also in the plot: the intercepts do not in average seem to be much different for the two treatments, while bprs seems to decrease a bit faster with the first treatment but the difference is not very large. Also, some of the slopes are positive and not negative.

I suppose replicating Section 9.4 considering the BtB data is not necessary as we already did the similar things with the BPRS data. 

