# Insert chapter 2 title here

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

#### Task 1.
We consider a cross-sectional dataset from 2014 containing statistics students' answers to questions regarding teaching and learning. The data contains 7 variables: gender (Male or Female), age, attitude (related to learning statistics, scale 1-5), deep (average of answers related to deep learning, scale 1-5), stra (average of answers related to strategic learning, scale 1-5), surf (average of answers related to surface learning), and points (supposedly exam points). Besides the variable gender, all the variables are numeric. The data is loaded below where also its structure and dimensions are pre-examined. 

```{r}
learning2014 <- read.csv("data/learning2014.csv")
head(learning2014, n=5) # The first 5 rows of the data table
str(learning2014) # Variables and their types
```
